# 11.21.2021

# This code is used for analyzing the SHAPE data. 

# Load libraries
```{r setup}
# This will set the working directory 
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
library(here)
here::i_am('key')
# knitr::opts_knit$set(root.dir = here())
library(readxl)
library(tidyverse)
library(plyr)
library(patchwork)
library(ggthemes)
library(viridis)
library(ggplot2)
library(data.table)
library(magrittr)
library(yangR)
library(viridis)
library(lme4)
library(ggthemes)
```

# Plotting functions:
```{r}
formatbeta = function (beta) 
{
    formatC(round(beta, 3), 3, format = "f")
}

formatpval = function (pval) 
{
    ifelse(pval > 0.001, format(round(pval, 2), nsmall = 2), 
        formatC(pval, format = "e", digits = 2))
}

plot_lm = function(data, x, y){
  require(yangR)
  require(magrittr)
  require(ggthemes)
  lm(as.formula(paste0(y , '~', x)), data) %>% summary %>% coef -> coefs
  beta = formatbeta(coefs[2,1])
  p = formatpval(coefs[2,4])
  ggplot(data, aes_string(x, y)) + geom_point() + geom_smooth(method = 'lm') + ggtitle(paste0('Effect estimate = ', beta, ', p-value = ', p)) + theme_classic()
}

plot_box = function(data, x, y){
  require(yangR)
  require(magrittr)
  require(ggthemes)
  t.test(as.formula(paste0(x , '~', y)), data = data, na.action = na.exclude) -> coefs
  beta = formatbeta(diff(coefs$estimate))
  p = formatpval(coefs$p.value)
  ggplot(data, aes_string(y, x)) + geom_boxplot() + geom_point() + ggtitle(paste0('Effect estimate = ', beta, ', p-value = ', p)) + theme_classic() + stat_summary(fun.data = give.n, geom = 'text')
}
```

# Read in data:
```{r}
alb1 = read_excel("/Volumes/JHPCE/dcs01/static/mito/SHAPE/qPCR/Monochrome/01122021_SHAPE1ALB.xlsx", sheet = "Results", skip = 37)
dlp1 = read_excel("/Volumes/JHPCE/dcs01/static/mito/SHAPE/qPCR/Monochrome/01122021_SHAPE1DLP.xlsx", sheet = "Results", skip = 37)

  alb1 = alb1[-(385:389),]
  dlp1 = dlp1[-(385:389),]
  alb1$Alb_CT = as.numeric(alb1$CT)
  dlp1$Dlp_CT = as.numeric(dlp1$CT)
  df = merge(dlp1, alb1, by = c('Well Position', 'Sample Name'))

  colnames(df) = make.names(colnames(df))
  
  df$Row = substr(df$`Well.Position`, 1, 1)
  df$Col = as.numeric(substr(df$`Well.Position`, 2, 4))
  df$deltaCT = df$Alb_CT - df$Dlp_CT
  
  df$Original.Plate = 0
  
  df$Original.Plate[which(df$Row %in% c('A', 'C', 'E', 'G', 'I', 'K', 'M', 'O') & df$Col %in% 1:12)] = 1
  df$Original.Plate[which(df$Row %in% c('B', 'D', 'F', 'H', 'J', 'L', 'N', 'P') & df$Col %in% 1:12)] = 2
  df$Original.Plate[which(df$Row %in% c('A', 'C', 'E', 'G', 'I', 'K', 'M', 'O') & df$Col %in% 13:24)] = 3
  df$Original.Plate[which(df$Row %in% c('B', 'D', 'F', 'H', 'J', 'L', 'N', 'P') & df$Col %in% 13:24)] = 4

  p1 = subset(df, Original.Plate == 1)
p2 = subset(df, Original.Plate == 2)
p3 = subset(df, Original.Plate == 3)
p4 = subset(df, Original.Plate == 4)

p1$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)
p2$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)
p3$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)
p4$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)

p1$Old.Col = as.numeric(p1$Col)
p2$Old.Col = as.numeric(p2$Col)
p3$Old.Col = as.numeric(p3$Col) - 12
p4$Old.Col = as.numeric(p4$Col) - 12

all = rbind(p1, p2, p3, p4)
all$Old.Well = paste0(all$Old.Row, all$Old.Col)
```

## Add Sample Names to data
### Plate 1
```{r}
key = as.data.frame(read_excel("/Volumes/JHPCE/dcs01/static/mito/SHAPE/qPCR/Monochrome/SHAPE Biomek 4000 Final Excel Sheet.1-13-2021.xlsx"))
colnames(key) = make.names(colnames(key))

mono.plate1 = subset(key, X96.well.Dilution.Plate.Name %in% c(paste0('SHAPEDilutionPlate', 1:4)))
mono.plate1$Original.Plate = as.numeric(substr(mono.plate1$X96.well.Dilution.Plate.Name, 19, 20))

w.key = merge(all, mono.plate1, by.x = c('Original.Plate', 'Old.Well'), by.y = c('Original.Plate', 'well.Location.in.96.well.Dilution.plate..Random.')) # all good.
w.key.1 = w.key
```

### Plate 2
```{r}

alb1 = read_excel("/Volumes/JHPCE/dcs01/static/mito/SHAPE/qPCR/Monochrome/01122021_SHAPE2_ALB.xlsx", sheet = "Results", skip = 37)
dlp1 = read_excel("/Volumes/JHPCE/dcs01/static/mito/SHAPE/qPCR/Monochrome/01122021_SHAPE2_DLP.xlsx", sheet = "Results", skip = 37)

alb1 = alb1[-(385:389),]
dlp1 = dlp1[-(385:389),]
alb1$Alb_CT = as.numeric(alb1$CT)
dlp1$Dlp_CT = as.numeric(dlp1$CT)
df = merge(dlp1, alb1, by = c('Well Position', 'Sample Name'))

colnames(df) = make.names(colnames(df))

df$Row = substr(df$`Well.Position`, 1, 1)
df$Col = as.numeric(substr(df$`Well.Position`, 2, 4))
df$deltaCT = df$Alb_CT - df$Dlp_CT

df$Original.Plate = 0

df$Original.Plate[which(df$Row %in% c('A', 'C', 'E', 'G', 'I', 'K', 'M', 'O') & df$Col %in% 1:12)] = 1
df$Original.Plate[which(df$Row %in% c('B', 'D', 'F', 'H', 'J', 'L', 'N', 'P') & df$Col %in% 1:12)] = 2
df$Original.Plate[which(df$Row %in% c('A', 'C', 'E', 'G', 'I', 'K', 'M', 'O') & df$Col %in% 13:24)] = 3
df$Original.Plate[which(df$Row %in% c('B', 'D', 'F', 'H', 'J', 'L', 'N', 'P') & df$Col %in% 13:24)] = 4

p1 = subset(df, Original.Plate == 1)
p2 = subset(df, Original.Plate == 2)
p3 = subset(df, Original.Plate == 3)
p4 = subset(df, Original.Plate == 4)

p1$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)
p2$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)
p3$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)
p4$Old.Row = rep(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H'), each = 12)

p1$Old.Col = as.numeric(p1$Col)
p2$Old.Col = as.numeric(p2$Col)
p3$Old.Col = as.numeric(p3$Col) - 12
p4$Old.Col = as.numeric(p4$Col) - 12

all = rbind(p1, p2, p3, p4)
all$Old.Well = paste0(all$Old.Row, all$Old.Col)

# add sample names to data for plate 2
mono.plate2 = subset(key, X96.well.Dilution.Plate.Name %in% c(paste0('SHAPEDilutionPlate', 5:8)))
mono.plate2$Original.Plate = as.numeric(substr(mono.plate2$X96.well.Dilution.Plate.Name, 19, 20)) - 4
w.key = merge(all, mono.plate2, by.x = c('Original.Plate', 'Old.Well'), by.y = c('Original.Plate', 'well.Location.in.96.well.Dilution.plate..Random.')) # all good.
w.key.2 = w.key
```

# Combine 1 + 2
```{r}
w.key.1.trim = dplyr::select(w.key.1, Subject, Visit, deltaCT)
w.key.2.trim = dplyr::select(w.key.2, Subject, Visit, deltaCT)
w.key.1.trim$Monochrome.Plate = 1
w.key.2.trim$Monochrome.Plate = 2
all.shape.data = rbind(w.key.1.trim, w.key.2.trim)
saveRDS(all.shape.data, file = '/Volumes/JHPCE/SHAPE/R_objects/all.shape.data.rds')
```
  
# QC starts
```{r}
all.shape.data = readRDS('R_objects/all.shape.data.rds')
```

## Remove deltaCT too low:
```{r}
fail_mono = subset(all.shape.data, deltaCT < 7)

# table(fail_mono$Subject)
# View(subset(all.shape.data, Subject %in% c(2001, 2039, 2066, 2087, 2195, 3110, 5091, 5104)))
```

## Remove genetic fingerprinting failures:
```{r}
bad1 = readRDS('R_objects/bad_samples1.rds')
bad2 = readRDS('R_objects/bad_samples2.rds')
bad3 = readRDS('R_objects/bad_samples3.rds')
bad1 = substr(bad1, 1, 4)

bad = unique(c(bad1, bad2$Subject))
```

## Remove bad data
```{r}
bad.data = subset(all.shape.data, deltaCT < 7)
no.bad.data = subset(all.shape.data, deltaCT >= 7)

no.bad.data.sequenom = subset(no.bad.data, Subject %!in% bad)
no.bad.data.sequenom$Subj.Vis = paste(no.bad.data.sequenom$Subject, no.bad.data.sequenom$Visit)
no.bad.data.sequenom2 = subset(no.bad.data.sequenom, Subj.Vis %!in% bad3)

library(lme4)
lmer(deltaCT ~ (1|Monochrome.Plate), no.bad.data.sequenom2) %>% summary

no.bad.data.sequenom2$deltaCT_PlateAdjust <- scale(resid(lmer(deltaCT ~ (1|Monochrome.Plate), no.bad.data.sequenom2)))

ggplot(no.bad.data.sequenom2, aes(deltaCT, col = as.factor(Monochrome.Plate))) + geom_density()
ggplot(no.bad.data.sequenom2, aes(deltaCT_PlateAdjust, col = as.factor(Monochrome.Plate))) + geom_density()
```

## Adjust for plate effects
```{r}
no.bad.data.sequenom = no.bad.data.sequenom2
w.key.1.trim.bl = subset(no.bad.data.sequenom, Visit == 'Baseline')
w.key.1.trim.fi = subset(no.bad.data.sequenom, Visit == 'Final')
quicklook = merge(w.key.1.trim.bl, w.key.1.trim.fi, by = 'Subject', suffixes = c('_BL', '_Final'), all.x = T, all.y = T)

quicklook$Study = as.factor(substr(quicklook$Subject, 1, 1))
```

## Save cleaned data
```{r}
# saveRDS(quicklook, file = 'R_objects/data.cleaned.rds')
```

# Make Table 1:
```{r}
library(htmlTable)
show = data.frame(Study = c('SHAPE3', 'SHAPE5'), N = c('77', '77'), Comorbidities = c('Overweight/obese + prediabetes/diabetes', 'Obese, otherwise healthy'), Group1 = c('Diet', 'Exercise + Low CHO'), Group2 = c('Diet + Exercise', 'Exercise + Low Fat'), Completed = c(55, 60), `Drop Out/Withdrew` = c(22, 17))
library(openxlsx)
# write.xlsx(show, file = '~/Dropbox/SHAPE_Paper/Table1.xlsx')

pheno = as.data.frame(read.table('/Volumes/JHPCE/dcs01/static/mito/SHAPE/phenotypes/new_phenotypes_3.26.18/phenoshapetotal.txt', header = T, sep = '\t'))
pheno %>% mutate(Study_Status = paste(Study, Study.Status)) %>% filter(Study != 'Shape2') -> filtered
table(filtered$Study_Status)
```

# Add phenotypes, create mtDNA-CN change variable
```{r}
shape.data = readRDS('../R_objects/data.cleaned.rds')
pheno = as.data.frame(read.table('../../../../static/mito/SHAPE/phenotypes/new_phenotypes_3.26.18/phenoshapetotal.txt', header = T, sep = '\t'))
shape.data$Study = paste0('Shape', shape.data$Study)
w.pheno = merge(shape.data, pheno, by.x = c('Subject', 'Study'), by.y = c('ID', 'Study'))
table(w.pheno$Study)
table(w.pheno$Study.Status)
w.pheno$Final_minus_BL = w.pheno$deltaCT_PlateAdjust_Final - w.pheno$deltaCT_PlateAdjust_BL
```


## Remove SHAPE2 and outlier individual
```{r}
w.pheno = subset(w.pheno, Study != 'Shape2')
# remove outlier individual
w.pheno = w.pheno[-unique(find.outliers(w.pheno$deltaCT_BL), find.outliers(w.pheno$deltaCT_Final)),]

w.pheno$Sex = ifelse(w.pheno$Male == 1, "Male", "Female")

saveRDS(w.pheno, file = '../final_shape3shape5.rds')
```

# Load dataframe:
```{r}
w.pheno = readRDS('../final_shape3shape5.rds')
```


# Examine known associations
## Age and sex
```{r}
sexplot = plot_box(w.pheno, 'deltaCT_PlateAdjust_BL', 'Sex') + ylab('Baseline mtDNA-CN')

library(splines)
lm(deltaCT_PlateAdjust_BL ~ Age, data = w.pheno) %>% summary
lm(deltaCT_PlateAdjust_BL ~ ns(Age, 2), data = w.pheno) %>% summary
ns(w.pheno$Age, 2)

m1 = lm(deltaCT_PlateAdjust_BL ~ Sex, data = w.pheno)
m2 = lm(deltaCT_PlateAdjust_BL ~ Sex + Age, data = w.pheno)

m3.5 = lm(deltaCT_PlateAdjust_BL ~ ns(Age, 2), data = w.pheno)
m3 = lm(deltaCT_PlateAdjust_BL ~ Sex + ns(Age, 2), data = w.pheno)

anova(m2, m3)

install.packages('lmtest')
library(lmtest)
A = logLik(m1)
B = logLik(m2)
(teststat <- -2 * (as.numeric(A)-as.numeric(B)))
(p.val <- pchisq(teststat, df = 1, lower.tail = FALSE))

A = logLik(m3.5)
B = logLik(m3)
(teststat <- -2 * (as.numeric(A)-as.numeric(B)))
(p.val <- pchisq(teststat, df = 1, lower.tail = FALSE))


ageplot = ggplot(w.pheno, aes(Age, deltaCT_PlateAdjust_BL)) + geom_point() + geom_smooth(method = lm, formula = y ~ ns(x, 2)) + ylab('Baseline mtDNA-CN') + theme_classic() + ggtitle('Spline at 52.6 years')

# ageplot = plot_lm(w.pheno, 'Age', 'deltaCT_PlateAdjust_BL')
# plot_lm(subset(w.pheno, Age >= 50), 'Age', 'deltaCT_PlateAdjust_BL')

library(patchwork)
sexplot + (ageplot + theme(axis.title.y=element_blank()))

length(na.omit(w.pheno$deltaCT_PlateAdjust_BL))
# saveRDS(w.pheno, 'R_objects/final_shape35_dataset.rds')
```

## Stratify by study
```{r}
ggplot(subset(w.pheno, Study == 'Shape3'), aes(Age, deltaCT_PlateAdjust_BL)) + geom_point() +     geom_smooth(method = lm, formula = y ~ ns(x, 2)) + ylab('Baseline mtDNA-CN') + theme_classic() + ggtitle('SHAPE3')
ggplot(subset(w.pheno, Study == 'Shape5'), aes(Age, deltaCT_PlateAdjust_BL)) + geom_point() +     geom_smooth(method = lm, formula = y ~ ns(x, 2)) + ylab('Baseline mtDNA-CN') + theme_classic() + ggtitle('SHAPE5')
lm(deltaCT_PlateAdjust_BL ~ ns(Age, 2), data = subset(w.pheno, Study == 'Shape5')) %>% summary
lm(deltaCT_PlateAdjust_BL ~ ns(Age, 2), data = subset(w.pheno, Study == 'Shape3')) %>% summary

sex3 = plot_box(subset(w.pheno, Study == 'Shape3'), 'deltaCT_PlateAdjust_BL', 'Sex') + ylab('Baseline mtDNA-CN') + ggtitle('')
sex5 = plot_box(subset(w.pheno, Study == 'Shape5'), 'deltaCT_PlateAdjust_BL', 'Sex') + ylab('Baseline mtDNA-CN') +ggtitle('')

age3 = ggplot(subset(w.pheno, Study == 'Shape3'), aes(Age, deltaCT_PlateAdjust_BL)) + geom_point() + geom_smooth(method = lm, formula = y ~ ns(x, 2)) + ylab('Baseline mtDNA-CN') + theme_classic()

age5 = ggplot(subset(w.pheno, Study == 'Shape5'), aes(Age, deltaCT_PlateAdjust_BL)) + geom_point() + geom_smooth(method = lm, formula = y ~ ns(x, 2)) + ylab('Baseline mtDNA-CN') + theme_classic()
(sex3+sex5)/(age3+age5)
```

# Correlation between BL and Final
```{r}
cor(w.pheno$deltaCT_PlateAdjust_BL, w.pheno$deltaCT_Final, use = 'complete.obs')
lm(deltaCT_PlateAdjust_BL~deltaCT_Final, data = w.pheno) %>% summary
# Make a plot showing correlations between baseline and final!
which(complete.cases(dplyr::select(w.pheno, deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final)) == T) %>% length

p = ggplot(w.pheno, aes(deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final, col = `Study`)) + geom_point() + ggtitle('Pearson correlation = 0.578') + xlab('Baseline mtDNA-CN') + ylab('Final mtDNA-CN') + geom_smooth(method = 'lm') + scale_color_viridis_d() + theme_classic()
p
```

## By study
```{r}
plot_box(w.pheno, 'deltaCT_PlateAdjust_BL', 'Study') + ylab('Baseline mtDNA-CN')
```

# Dropouts 
```{r}
w.pheno$Completed = ifelse(w.pheno$Completed == 0, 'No', 'Yes')
w.pheno$Completed = as.factor(w.pheno$Completed)

cor(as.numeric(w.pheno$Completed2), w.pheno$deltaCT_PlateAdjust_BL, use = 'complete.obs')

plot_box(w.pheno, 'deltaCT_PlateAdjust_BL', 'Completed') + xlab('Completed') + ylab('Baseline mtDNA-CN')

shape3 = plot_box(subset(w.pheno, Study == 'Shape3'), 'deltaCT_PlateAdjust_BL', 'Completed') + xlab('Completed') + ylab('Baseline mtDNA-CN')

shape5 = plot_box(subset(w.pheno, Study == 'Shape5'), 'deltaCT_PlateAdjust_BL', 'Completed') + xlab('Completed') + ylab('Baseline mtDNA-CN')

# Supplemental Figure 5 
shape3 + shape5 

dropout_plot = plot_box(w.pheno, 'deltaCT_PlateAdjust_BL', 'Completed') + xlab('Completed') + ylab('Baseline mtDNA-CN')

t.test(w.pheno$deltaCT_PlateAdjust_BL ~ w.pheno$Completed)
```

## Study Status
```{r}
lm(deltaCT_PlateAdjust_BL ~ Study.Status, data = w.pheno) %>% summary

# Only one person withdrew, so vast majority is dropouts
```

## Test Study Dropout against all available variables
```{r}
w.pheno2 = w.pheno[, -c(6, 11)]
w.pheno2$ENDOAI_D = NULL
w.pheno2$ENDOAI_P = NULL

save = as.data.frame(matrix(nrow = 1, ncol = 3))
colnames(save) = c('VarName', 'estimate', 'p.value')
to_test = colnames(w.pheno2)
for(i in 1:length(to_test)){
  var = to_test[i]
  outcome = ifelse(w.pheno2$Completed == 1, TRUE, FALSE) 
  predict = w.pheno2[,i]
  if(length(levels(as.factor(predict))) > 1){
    fit = lm(outcome ~ predict)
    fit = glm(outcome ~ predict, family = 'binomial')
    tidyfit = broom::tidy(fit)
    if(tidyfit$p.value[2] < 0.05){
      print(var)
      print(tidyfit$p.value[2])
      save = rbind(save, c(var, tidyfit$estimate[2], tidyfit$p.value[2]))
    }
  }
}

save = na.omit(save)

# Take out variables that are for later -->
save_new = save[-c(union(grep('6', save$VarName), grep('_D', save$VarName))),]

save_new2 = save_new[-union(grep('Attend', save_new$VarName), grep('LOG', save_new$VarName)),]

save_3 = save_new2[c(1, 2, 3, 6, 7, 8, 9, 10, 14), ]
```

### Dropout Table
```{r}
save_3$VarName
save_3$estimate = formatbeta(as.numeric(save_3$estimate))
save_3$p.value = formatpval(as.numeric(save_3$p.value))

library(htmlTable)
htmlTable(save_3[-c(1),], rnames = rep('', nrow(save_3)))
```

### Correlation Plot
```{r}
pairs(~Completed + deltaCT_PlateAdjust_BL + Age + SLEEPTimeWorkday0 + IVRT_0 + INF_Gamma.MSD_0, data = w.pheno2, lower.panel=twolines, diag.panel=mydiag.panel, upper.panel=panel.cor, label.pos=0.5, main="Correlation matrix", labels = c('Dropout', 'mtDNA', 'Age', 'Sleep', 'IVRT', 'IFN Gamma'))

pairs(~Completed + deltaCT_PlateAdjust_BL + Resistance.Volume.Program.Half_1 + Resistance.Volume.Program.Half_2 + Age + SLEEPTimeWorkday0 + IVRT_0 + INF_Gamma.MSD_0, data = w.pheno2, lower.panel=twolines, diag.panel=mydiag.panel, upper.panel=panel.cor, label.pos=0.5, main="Correlation matrix", labels = c('Dropout', 'mtDNA', 'ResistanceVol1', 'ResistanceVol2', 'Age', 'Sleep', 'IVRT', 'IFN Gamma'))

```

### Dropouts and vo2max, insulin resistance
```{r}
plot_box(w.pheno, 'QUICKI0', 'Completed')
plot_box(w.pheno, 'ETTVO2kg0', 'Completed')
plot_lm(w.pheno, 'Exercise', 'Completed')

glm(Completed ~ Exercise,family=binomial(link='logit'),data=w.pheno) %>% summary

table(w.pheno$Completed, w.pheno$Exercise)
```

### Stratify by study
```{r}
s3 = plot_box(subset(w.pheno, Study == 'Shape3'), 'deltaCT_PlateAdjust_BL', 'Completed') + xlab('Completed') + ylab('Baseline mtDNA-CN') 
s5 = plot_box(subset(w.pheno, Study == 'Shape5'), 'deltaCT_PlateAdjust_BL', 'Completed') + xlab('Completed') + ylab('Baseline mtDNA-CN') 
s3 + s5
```

# Insulin resistance
```{r}
w.pheno$QUICKI0_diabetes = resid(lm(QUICKI0 ~ as.factor(Diabetes), data = w.pheno, na.action = na.exclude))
w.pheno$QUICKI6_diabetes = resid(lm(QUICKI6 ~ as.factor(Diabetes), data = w.pheno, na.action = na.exclude))

ins_mt = plot_lm(w.pheno, 'deltaCT_PlateAdjust_BL', 'QUICKI0_diabetes') + xlab('Baseline mtDNA-CN') + ylab('Insulin sensitivity (adjusted for diabetes)')
ins_mt_final = plot_lm(w.pheno, 'deltaCT_PlateAdjust_Final', 'QUICKI6_diabetes') + xlab('6 month mtDNA-CN') + ylab('6 month insulin sensitivity (adjusted for diabetes)')
ins_mt + ins_mt_final

# linear mixed model
for_mix = dplyr::select(w.pheno, Subject, Study, deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final, QUICKI0, QUICKI6, Diabetes)

for_mix %>% dplyr::select(Subject, Study, Diabetes, QUICKI0, QUICKI6) %>% gather(Visit, Sensitivity, c(QUICKI0, QUICKI6)) -> ins_gather

for_mix %>% dplyr::select(Subject, Study, Diabetes, deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final) %>% gather(Visit, mtDNA_CN, c(deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final)) -> mt_gather

mt_gather$Visit = ifelse(mt_gather$Visit == 'deltaCT_PlateAdjust_Final', 'Final', 'BL')
                                            
ins_gather$Visit = ifelse(ins_gather$Visit == 'QUICKI0', 'BL', 'Final')
mt_gather$Visit = ifelse(mt_gather$Visit == 'deltaCT_PlateAdjust_BL', 'BL', 'Final')

full_gather = merge(ins_gather, mt_gather)

ggplot(full_gather, aes(mtDNA_CN, Sensitivity, col = Visit)) + geom_point() + ylab("Insulin Sensitivity")

full_gather$Sensitivity_adj = resid(lm(Sensitivity ~ Diabetes, data = full_gather, na.action = na.exclude))

library(lme4)
mod = lmer(Sensitivity_adj ~ mtDNA_CN + (1|Visit), data = full_gather) 
full_gather$Sens_predict = predict(mod, na.action = na.exclude)

# p-val from t-value:
pt(-3.217795, df = 248)

try2 = ggplot(full_gather, aes(mtDNA_CN, Sensitivity_adj, col = Visit)) + geom_point() + ylab("Insulin Sensitivity (adjusted for Diabetes status)") + scale_color_viridis_d() + theme_classic() + geom_line(data = full_gather, aes(mtDNA_CN, Sens_predict)) + ggtitle('P-value for mixed linear model: 7.32e-4')
```

# Baseline vs. Change
```{r}
w.pheno$mtDNA_change = w.pheno$deltaCT_PlateAdjust_Final - w.pheno$deltaCT_PlateAdjust_BL

plot_lm(w.pheno, 'deltaCT_PlateAdjust_BL', 'mtDNA_change') + xlab('Baseline mtDNA-CN') + ylab('Change in mtDNA-CN after 6 months')

# show this better:
# plot_lm(w.pheno, 'deltaCT_PlateAdjust_BL', 'deltaCT_PlateAdjust_Final') + xlab('Baseline mtDNA-CN') + ylab('Final mtDNA-CN ')

# if you have a negative baseline, you're more likely to increase?

# if you have a positive baseline, you're more likely to decrease?

w.pheno$BL_bin = ifelse(w.pheno$deltaCT_PlateAdjust_BL < 0, 'BL_negative', 'BL_positive')
w.pheno$change_bin = ifelse(w.pheno$mtDNA_change < 0, 'change_negative', 'change_positive')

table(w.pheno$BL_bin, w.pheno$change_bin)
```

## Absolute Value
```{r}
w.pheno$Abs_BL = abs(w.pheno$deltaCT_PlateAdjust_BL)
w.pheno$Abs_change = abs(w.pheno$mtDNA_change)

ggplot(subset(w.pheno, change_bin %in% c('change_negative', 'change_positive')), aes(Abs_BL, Abs_change, col = change_bin)) + geom_point() + xlab('Absolute value of baseline mtDNA-CN') + ylab('Absolute value of 6-month change in mtDNA-CN') + theme_classic()

w.pheno$base_times_change = w.pheno$mtDNA_change * w.pheno$deltaCT_PlateAdjust_BL

w.pheno$change_bin2 = ifelse(w.pheno$base_times_change < 0, 'Towards mean', 'Away from mean')

cb1 = ggplot(subset(w.pheno, change_bin %in% c('change_negative', 'change_positive')), aes(Abs_BL, Abs_change, col = change_bin)) + geom_point() + xlab('Absolute value of baseline mtDNA-CN') + ylab('Absolute value of 6-month change in mtDNA-CN') + theme_classic()

# Supplementary Figure XY
table(w.pheno$change_bin2)
w.pheno$change_bin2 = factor(w.pheno$change_bin2, levels = c('Towards mean', 'Away from mean'))
cb2 = ggplot(subset(w.pheno, change_bin %in% c('change_negative', 'change_positive')), aes(Abs_BL, Abs_change, col = change_bin2)) + geom_point() + xlab('Absolute value of baseline mtDNA-CN') + ylab('Absolute value of 6-month change in mtDNA-CN') + theme_classic() + scale_color_viridis(discrete = TRUE, name = 'Direction of change', labels = c('Towards mean (67)', 'Away from mean (38)')) + geom_smooth(method = 'lm', se = F)

(obs - exp)^2/(exp)
chisq = (67 -52.5)^2/(52.5) + (38-52.5)^2/(52.5)
pchisq(chisq, 1)
pchisq(chisq, 1, lower.tail = FALSE)
# Significant 0.004
check = subset(w.pheno, change_bin2 == 'Towards mean')

plot_lm(w.pheno, 'Abs_BL', 'Abs_change')
```

# Exercise sessions
```{r}
w.pheno$Exercise.Sessions.Attended2 = ifelse(w.pheno$Completed == 1, w.pheno$Exercise.Sessions.Attended, 0)

w.pheno$Exercise.Sessions.Attended2[which(is.na(w.pheno$Exercise.Sessions.Attended2))] = 0

w.pheno$mtDNA_change = w.pheno$deltaCT_PlateAdjust_Final - w.pheno$deltaCT_PlateAdjust_BL
w.pheno$Exercise = ifelse(w.pheno$Group %in% c('Diet'), 'No', 'Yes')

w.pheno$mtDNA_change_adjust_BL = resid(lm(mtDNA_change ~ deltaCT_PlateAdjust_BL, data = w.pheno, na.action = na.exclude))

sessionplot = plot_lm(w.pheno, 'Exercise.Sessions.Attended2', 'mtDNA_change_adjust_BL') + xlab('Number of exercise sessions attended') + ylab('6-month change in mtDNA-CN')
```

## Exclude diabetic individuals
```{r}
plot_lm(subset(w.pheno, Diabetes == 'Yes'), 'Exercise.Sessions.Attended2', 'mtDNA_change_adjust_BL') + xlab('Number of exercise sessions attended') + ylab('Change in mtDNA-CN after 6 months')

plot_lm(subset(w.pheno, Diabetes == 'No'), 'Exercise.Sessions.Attended2', 'mtDNA_change_adjust_BL') + xlab('Number of exercise sessions attended') + ylab('Change in mtDNA-CN after 6 months')

# Code people who didn't exercise as zero??
w.pheno$Exercise.Sessions.Attended2 = ifelse(w.pheno$Completed == 1, w.pheno$Exercise.Sessions.Attended, 0)
w.pheno$Exercise.Sessions.Attended2 = ifelse(w.pheno$Group == 'Diet', 0, w.pheno$Exercise.Sessions.Attended)

plot_lm(w.pheno, 'Exercise.Sessions.Attended2', 'Final_minus_BL') + xlab('Number of exercise sessions attended') + ylab('Change in mtDNA-CN after 6 months')
```

### Stratify by study
```{r}
plot_lm(subset(w.pheno, Study == 'Shape3'), 'Exercise.Sessions.Attended', 'Final_minus_BL') + xlab('Number of exercise sessions attended') + ylab('Change in mtDNA-CN after 6 months')

plot_lm(subset(w.pheno, Study == 'Shape5'), 'Exercise.Sessions.Attended', 'Final_minus_BL') + xlab('Number of exercise sessions attended') + ylab('Change in mtDNA-CN after 6 months')
```

## Binary exercise
```{r}
explot = plot_box(w.pheno, 'mtDNA_change_adjust_BL', 'Exercise') + xlab('Exercise') + ylab('6-month change in mtDNA-CN') 

explot + (sessionplot + theme(axis.title.y=element_blank()))

lm(mtDNA_change ~ ETTV02kg_D + ETTVO2kg0 +  deltaCT_PlateAdjust_BL, data = w.pheno) %>% summary
t.test(mtDNA_change ~ Exercise, data = w.pheno)
# no significant difference
```

# mtDNA-CN and diabetes
```{r}
m1 = lm(mtDNA_change ~ Exercise, data = w.pheno)
m2 = lm(mtDNA_change ~ Exercise * Diabetes, data = w.pheno)
anova(m1, m2)
```

## vo2max

#### baseline/final mixed
```{r}
for_mix = dplyr::select(w.pheno, Subject, Study, deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final, ETTVO2kg0, ETTVO2kg6, Diabetes, Age, Sex)

for_mix %>% dplyr::select(Subject, Study, Diabetes, deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final) %>% gather(Visit, mtDNA_CN, c(deltaCT_PlateAdjust_BL, deltaCT_PlateAdjust_Final)) -> mt_gather

for_mix %>% dplyr::select(Subject, Study, Diabetes, ETTVO2kg0, ETTVO2kg6, Age, Sex) %>% gather(Visit, VO2max, c(ETTVO2kg0, ETTVO2kg6)) -> max_gather

max_gather$Visit = ifelse(max_gather$Visit == 'ETTVO2kg0', 'BL', 'Final')

max_mt = merge(max_gather, mt_gather)

ggplot(max_mt, aes(mtDNA_CN, VO2max, col = Visit)) + geom_point() + ylab("VO2max")

library(lme4)
mod = lmer(VO2max ~ mtDNA_CN + ns(Age, 2) + Sex + Visit + (1|Subject) + Study, data = max_mt) 

max_mt$VO2_predict = predict(mod, na.action = na.exclude)
summary(mod)
mod %>% summary %>% coef
# p-val from t-value:
pt(-0.1384982, df = 244)

ggplot(max_mt, aes(mtDNA_CN, VO2max, col = Visit)) + geom_point() + ylab("VO2max") + scale_color_viridis_d() + theme_classic() + geom_line(data = max_mt, aes(mtDNA_CN, VO2_predict)) + ggtitle('P-value for mixed linear model: 0.44') + xlab('mtDNA-CN')

lm(ETTV02kg_D ~ ETTVO2kg0, data = w.pheno) %>% summary

plot_lm(w.pheno, 'ETTVO2kg0', 'ETTV02kg_D')
plot_lm(w.pheno, 'mtDNA_change_adjust_BL', 'ETTV02kg_D') + xlab('6-month change in mtDNA-CN') + ylab('6-month change in VO2max')
```

```{r}
ggplot(w.pheno, aes(Exercise.Sessions.Attended, ETTV02kg_D, col = as.factor(Diabetic))) + geom_point()
ggplot(w.pheno, aes(Exercise.Sessions.Attended, ETTV02kg_D, col = as.factor(Sex))) + geom_point()

# Supplemental Figure 1
vo2max_lm = plot_lm(w.pheno, 'Exercise.Sessions.Attended2', 'ETTV02kg_D') + ylab('Change in VO2max over 6 months') + xlab('Exercise Sessions Attended')

cor(w.pheno$Exercise.Sessions.Attended2, w.pheno$ETTV02kg_D, use = 'complete.obs')

# Supplemental Figure 2
bmi_lm = plot_lm(w.pheno, 'Exercise.Sessions.Attended2', 'ANTHROBMI_D') + ylab('Change in BMI over 6 months') + xlab('Exercise Sessions Attended')

cor(w.pheno$Exercise.Sessions.Attended2, w.pheno$ANTHROBMI_D, use = 'complete.obs')


vo2max_box = plot_box(w.pheno, 'ETTV02kg_D', 'Exercise') + ylab('Change in VO2max over 6 months')

vo2max_box + vo2max_lm 


bl_vo2max = plot_lm(w.pheno, 'ETTVO2kg0', 'deltaCT_PlateAdjust_BL') + xlab('Baseline VO2max') + ylab('Baseline mtDNA-CN') 

final_vo2max = plot_lm(w.pheno, 'ETTVO2kg6', 'deltaCT_PlateAdjust_Final') + xlab('Final VO2max') + ylab('Final mtDNA-CN') 

diff_vo2max = plot_lm(w.pheno, 'ETTV02kg_D', 'mtDNA_change') + xlab('6-month change in VO2max') + ylab('6-month change in mtDNA-CN') 

bl_vo2max + final_vo2max
```


# change in mt
## age sex
```{r}
plot_lm(w.pheno, 'Age', 'mtDNA_change')
plot_box(w.pheno, 'mtDNA_change', 'Sex')
```


# why bimodal distribution for non-exercisers?
```{r}
no.shape2 = w.pheno
no.exercisers = subset(no.shape2, Exercise == 'No')
no.exercisers$Diabetes %>% table
ggplot(no.exercisers, aes(mtDNA_change, col = Diabetes)) + geom_density() + geom_rug()
ggplot(no.exercisers, aes(mtDNA_change, col = Gender)) + geom_density() + geom_rug()
ggplot(no.exercisers, aes(mtDNA_change, col = Ethnicity)) + geom_density() + geom_rug()
table(noshape2$Ethnicity)

```

# change in vo2max
```{r}
no.shape2$VO2max_change = no.shape2$ETTVO2kg6 - no.shape2$ETTVO2kg0

lm(VO2max_change ~ Percent.Exercise.Sessions.Attended, data = subset(no.shape2, Exercise == 'Yes')) %>% summary
lm(mtDNA_change ~ VO2max_change, data = no.shape2) %>% summary
lm(mtDNA_change ~ VO2max_change, data = subset(no.shape2, Diabetes == 'Yes')) %>% summary
lm(mtDNA_change ~ VO2max_change, data = subset(no.shape2, Diabetes == 'No')) %>% summary

ggplot(no.shape2, aes(VO2max_change, mtDNA_change, col = Diabetes)) + geom_point() + geom_smooth(method = 'lm')

ggplot(no.shape2, aes(VO2max_change, mtDNA_change, col = Exercise)) + geom_point() + geom_smooth(method = 'lm')

```

# Load
```{r}
getwd()
# saveRDS(w.pheno, file = 'pheno.rds')

w.pheno = readRDS('pheno.rds')
```

# Change into format where can use mixed models:
```{r}
# Need: Age, Sex, RMR, WHR, DXALEANMASS, MetSyn, QUICKI, Gluc, HBA1c, BMI
baseline = dplyr::select(w.pheno, Subject, Study, Age, Sex, deltaCT_PlateAdjust_BL, RMR_Estimated_0, ANTHROWHR0, DXALeanMass0, MSMetsyn0, QUICKI0, Gluc0, HA1C0, ANTHROBMI0, Diabetes, ETTVO2kg0, Diabetic) %>% mutate(mtDNA_CN = deltaCT_PlateAdjust_BL, Visit = "Baseline", RMR = RMR_Estimated_0, WHR = ANTHROWHR0, DXALeanMass = DXALeanMass0, MSMetsyn = MSMetsyn0, QUICKI = QUICKI0, Gluc = Gluc0, HbA1c = HA1C0, BMI = ANTHROBMI0, ETTVO2kg = ETTVO2kg0, Diabetic = Diabetic)
baseline2 = na.omit(baseline)

final = dplyr::select(w.pheno, Subject, Study, Age, Sex, deltaCT_PlateAdjust_Final, RMR_Estimated_6, ANTHROWHR6, DXALeanMass6, MSMetsyn6, QUICKI6, Gluc6, HA1C6, ANTHROBMI6, Diabetes, ETTVO2kg6, Diabetic) %>% mutate(mtDNA_CN = deltaCT_PlateAdjust_Final, Visit = "Final", RMR = RMR_Estimated_6, WHR = ANTHROWHR6, DXALeanMass = DXALeanMass6, MSMetsyn = MSMetsyn6, QUICKI = QUICKI6, Gluc = Gluc6, HbA1c = HA1C6, BMI = ANTHROBMI6, ETTVO2kg = ETTVO2kg6, Diabetic = Diabetic)
final2 = na.omit(final)

bothvis = rbind.fill(baseline2, final2) %>% select(Subject, Study, Visit, Age, Sex, mtDNA_CN, RMR, WHR, DXALeanMass, MSMetsyn, QUICKI, Gluc, HbA1c, Diabetes, BMI, ETTVO2kg, Diabetic)
```

## save
```{r}
getwd()
saveRDS(bothvis, file = 'bothvis.rds')
```

# Mixed models:
```{r}
bothvis = readRDS(file = '../bothvis.rds')

library(lme4)
bothvis$Subject = as.factor(bothvis$Subject)
```

## function
```{r}
check = function(variable = 'DXALeanMass', bothvis){
  form = as.formula(paste0(variable, ' ~ mtDNA_CN + Age + as.factor(Sex) + Visit + (1|Subject)'))
  m = lmer(form, data = bothvis)
  coefs = coef(summary(m))
  return = (coefs['mtDNA_CN',])
  p = 2*pt(q=abs(as.numeric(return[3])), lower.tail=FALSE, df = 240)
  return_vector = c(variable, as.numeric(return[1]), as.numeric(return[2]), p)
  return(return_vector)
}
```

### do for all:
```{r}
dxa = check("DXALeanMass", bothvis)
insulin = check("QUICKI", bothvis)
rmr = check("RMR", bothvis)
gluc = check("Gluc", bothvis)

# MetSyn on its own: 
bothvis$MSMetsyn = as.factor(bothvis$MSMetsyn)

form = as.formula(paste0(variable, ' ~ mtDNA_CN + Age + as.factor(Sex) + Visit + (1|Subject)'))
m = glmer(form, data = bothvis, family = 'binomial')
coefs = coef(summary(m))
return = (coefs['mtDNA_CN',])
p = 2*pt(q=abs(as.numeric(return[3])), lower.tail=FALSE, df = 240)
metsyn = c(variable, as.numeric(return[1]), as.numeric(return[2]), p)

hb1ac = check("HbA1c", bothvis)
```

### make table
```{r}
library(htmlTable)

full = as.data.frame(rbind(dxa, insulin, rmr, gluc, metsyn, hb1ac))
colnames(full) = c('var', 'beta', 'std', 'p')

full$beta = as.numeric(full$beta)
full$std = as.numeric(full$std)
full$p = as.numeric(full$p)
full$p.adjust = p.adjust(full$p, method = 'fdr')

for_show = full
for_show$beta = formatbeta(for_show$beta)
for_show$std = formatbeta(for_show$std)
for_show$p = formatpval(for_show$p)
for_show$p.adjust = formatpval(for_show$p.adjust)

for_show$var = c('Muscle mass', 'Insulin sensitivity', 'Resting metabolic rate', 'Baseline glycemia', 'Metabolic syndrome', 'HbA1c')
htmlTable(for_show, rnames = rep('', nrow(full)), header = c('Secondary outcome', 'Effect size estimate', 'Standard error', 'P-value', 'FDR-adjusted P-value'))
```

### check quicki
```{r}
form = as.formula(paste0('QUICKI ~ mtDNA_CN + Age + as.factor(Sex) + Visit + (1|Subject) + Diabetic'))

m = lmer(form, data = bothvis)
coefs = coef(summary(m))
return = (coefs['mtDNA_CN',])
p = 2*pt(q=abs(as.numeric(return[3])), lower.tail=FALSE, df = 240)
return_vector = c('QUICKI', as.numeric(return[1]), as.numeric(return[2]), p)

# add drug info?
library(readxl)
meds = read_excel('../../../../static/mito/SHAPE/phenotypes/MedicationsS3_5.xlsx')
colnames(meds) = make.names(colnames(meds))
meds$MEDInsulin

# Only one individual on insulin
# You looked at statins and they weren't significant
```

## adjust for MetSyn
```{r}
check_metsyn = function(variable = 'DXALeanMass', bothvis){
  form = as.formula(paste0(variable, ' ~ mtDNA_CN + MSMetsyn + Age + as.factor(Sex) + Visit + (1|Subject)'))
  m = lmer(form, data = bothvis)
  coefs = coef(summary(m))
  return = (coefs['mtDNA_CN',])
  p = 2*pt(q=abs(as.numeric(return[3])), lower.tail=FALSE, df = 240)
  return_vector = c(variable, as.numeric(return[1]), as.numeric(return[2]), p)
  return(return_vector)
}
```

### run for all
```{r}
dxa = check_metsyn("DXALeanMass", bothvis)
insulin = check_metsyn("QUICKI", bothvis)
rmr = check_metsyn("RMR", bothvis)
gluc = check_metsyn("Gluc", bothvis)
hb1ac = check_metsyn("HbA1c", bothvis)
```

### make table:
```{r}
full = as.data.frame(rbind(dxa, insulin, rmr, gluc, hb1ac))
colnames(full) = c('var', 'beta', 'std', 'p')

full$beta = as.numeric(full$beta)
full$std = as.numeric(full$std)
full$p = as.numeric(full$p)
full$p.adjust = p.adjust(full$p, method = 'fdr')

for_show = full
for_show$beta = formatbeta(for_show$beta)
for_show$std = formatbeta(for_show$std)
for_show$p = formatpval(for_show$p)
for_show$p.adjust = formatpval(for_show$p.adjust)

for_show$var = c('Muscle mass', 'Insulin sensitivity', 'Resting metabolic rate', 'Baseline glycemia', 'HbA1c')
htmlTable(for_show, rnames = rep('', nrow(full)), header = c('Secondary outcome', 'Effect size estimate', 'Standard error', 'P-value', 'FDR-adjusted P-value'))
```


